A Plasma Torus Around a Young Low-Mass Star

### 2025-01-27 paper/ms.tex
update

\begin{abstract}
\normalfont
% v1: removed a sentence for wordcount.  v0 is under abstract_title.txt
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting clumps of opaque material corotating with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.
The composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these stars are
explained by complex distributions of dark starspots or bright
faculae distributed across their surfaces \cite{Koen2021}.  Here, we
present time-series spectroscopy and photometry of a \stagemyr\
million year old complex periodic variable (CPV), TIC~141146667. The
spectra show coherent sinusoidal Balmer emission at up to four times
the star's equatorial velocity, demonstrating the presence of
extended clumps of circumstellar plasma --- a plasma torus.  Given
that long-lived condensations of cool ($10^4$ K) plasma can persist
in the hot ($10^6$ K) coronae of stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that such condensations can become
optically thick around the lowest-mass stars, although the exact
source of opacity remains unclear.
\end{abstract}

@@

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been discovered by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
defined by their highly structured and periodic optical light curves, 
and most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features.  However, no independent evidence has
yet been acquired for the presence of circumstellar material in these
objects.  Since transiting circumstellar clumps would geometrically
imply an intrinsic occurrence a few to ten times the observed rate,
the question of whether circumstellar clumps exist in these systems
has the potential to be applicable to 10-30\% of M dwarfs during their
early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, caused by long-lived, corotating clumps of cool
plasma \cite{CollierCameron1989,Donati2000,Dunstone2006,Skelly2008}.
Such clumps are forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
magnetic field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs along the apexpoint furthest from the star
\cite{Waugh2022}.
However, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether these two areas of study have any direct connection.
Nonetheless, CPVs do respond to sudden magnetic field changes: the
otherwise long-lived eclipse features often disappear immediately
following stellar flares \cite{Stauffer2017,Bouma2024}.

@@

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  As expected
based on other CPVs \cite{Bouma2024}, the photometric shape of
TIC~141146667 evolved following the 2022 discovery
data, while nonetheless remaining complex.  In February 2024, the
average photometric signal showed a gradual brightening over 45\% of
the period, followed by a complex eclipse-like feature spanning 55\%
of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.
Its W-shape is suggestive of eclipse geometries seen in forward models
of warped plasma tori \cite{Townsend2008}.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
sharp transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to those
in Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.  The usual interpretation for
such variability comes from a loose analogy to
quiescent solar prominences and filaments, which are cool
condensations of plasma in the solar corona that can last days to
weeks \cite{VialEngvold2015}.  This plasma is called a prominence when
viewed in emission against the dark backdrop of space, and a filament
when viewed in absorption against the solar disk.  In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic or centrifugal force
capable of sustaining them.  However for stars with magnetospheric
radii $R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a plasma parcel can have a local minimum
outside $R_{\rm c}$, enabling the material to be sustained for
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

Our Keck/HIRES observations are the first high-resolution time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
Characteristic densities and masses of these
clumps are $n \sim 10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see
Supplementary Methods Section~\ref{subsec:model}), a similar density
to solar prominences, but ten to one hundred times more massive.  This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} since such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios
in which the circumstellar material is made only of dust are also
ruled out.  While dust may be present, to explain the H$\alpha$
emission the circumstellar clumps must include plasma with a
significant population of hydrogen atoms in the $n$=3 excited state.
While this plasma is undoubtedly sculpted by the star's magnetic
field, it could plausibly originate from three sites: the star, an old
and undetected disk, or outgassing rocky bodies.  This latter
possibility would render CPVs as extrasolar analogs of the Jupiter-Io
plasma torus (CITE).

@@

The material's composition -- either pure plasma, or else a dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical have
previously shown chromaticity consistent with dust \cite{Tanimoto2020,Gunther2022,Koen2023},
a gray opacity source such as electron scattering in a plasma
transiting over starspots could also produce chromatic features
\cite{Rackham2018}.  The composition and size distribution of any dust
that is present could be most easily resolved by measuring the
extinction curve for one or more CPVs from 1-10\,$\mu$m.  A
dust composition similar to debris from rocky bodies seen around white
dwarfs \cite{Reach2009} would indicate a rocky-body origin.  A 
composition closer to the ISM would be indicative of condensed dust in
an M dwarf wind, similar to that formed in the environments of more
evolved stars \cite{Marigo2008}.

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
(magneto)-hydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008} and to explore the effects of
non-dipolar field contributions.
In particular, dynamo simulations of fully-convective M dwarfs have
suggested the possibility that global-scale mean fields can be
confined to a single hemisphere \cite{Brown2020}; such fields would
yield accumulations of circumstellar material quite different
from those that have previously been explored \cite{Townsend2008}.
Observationally, spectropolarimetry has the potential to
assess both the field strength and topology.  A more direct probe
however might be to connected the recent work \cite{Kaur2024} showing
that CPVs are variable radio emitters, exhibiting emission components
that can be both persistent, as well as short-lived and highly
polarized.  This opens the prospects for detecting radio emission
from the electron cyclotron maser instability, which would
provide a direct measurement of the field strength at the site of the
emitting region.

@@

{\it Effective temperature, radius, and mass}
We adopt the results from the spectral energy distribution fitting
exercise described by \cite{Bouma2024}.
In brief, this approach used
\texttt{astroARIADNE}...
\cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012} assuming the
\cite{Asplund2009} solar abundances, and the
\cite{Barber2006} water line lists.  
We used broadband magnitudes from Gaia DR2, APASS, 2MASS, SDSS, and WISE
  $W1$ and $W2$.  (explain)...
%%%%%%%%
%\texttt{astroARIADNE} compares the measured broadband flux
%measurements against pre-computed model grids, and by default fits for
%six parameters: $\{ T_{\rm eff}, R_\star, A_{\rm V}, \log g, [{\rm
%Fe/H}], d \}$.  The distance  prior is drawn from
%\citet{2021AJ....161..147B}.  The surface gravity and metallicity are
%generally unconstrained.  Given our selection criteria for the stars,
%we assumed the following priors for the temperature, stellar size, and
%extinction:
%\begin{align}
%  T_{\rm eff} / {\rm K}    &\sim \mathcal{U}(2000, 8000), \\
%  R_\star / R_\odot  &\sim \mathcal{U}(0.1, 1.5), \\
%  A_{\rm V} / {\rm mag}    &\sim \mathcal{U}(0, 0.2),
%\end{align}
%for \deleted{$\mathcal{N}$ the Gaussian and }$\mathcal{U}$ the uniform
%distribution\deleted{s, and $\mathcal{T}_{\rm N}(\mu, \sigma, a, b)$ a
%truncated normal distribution with mean $\mu$, standard deviation
%$\sigma$, and lower and upper bounds $a$ and $b$}.  We validated our
%chosen upper bound on $A_{\rm V}$ using a 2MASS color-color diagram.
%Finally, using \texttt{Dynesty} \citep{2020MNRAS.493.3132S}, we
%sampled the posterior probability assuming the default Gaussian
%likelihood, and set a stopping threshold of ${\rm d}\log \mathcal{Z} <
%0.01$, where $\mathcal{Z}$ denotes the evidence.

>>>>

\begin{abstract}
\normalfont
% v1: removed a sentence for wordcount.  v0 is under abstract_title.txt
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting clumps of opaque material corotating with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
dark starspots or bright faculae distributed across the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a young ($t$=16\,Myr), rapidly-rotating
($P$=3.9\,hr) CPV, TIC~141146667. The spectra show coherent
sinusoidal Balmer emission at up to four times the star's equatorial
velocity, demonstrating the presence of extended clumps of
circumstellar plasma --- a plasma torus.  Given that long-lived
condensations of cool ($10^4$ K) plasma can persist in the hot
($10^6$ K) coronae of stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that such condensations can become
optically thick around the lowest-mass stars, although the exact
source of opacity remains unclear.
\end{abstract}

@@

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
defined by their highly structured and periodic optical light curves, 
and most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features.  However, no independent evidence has
yet been acquired for the presence of circumstellar material around these
objects.  Since transiting circumstellar clumps would geometrically
imply an intrinsic occurrence a few to ten times the observed rate,
the question of whether circumstellar clumps exist in these systems
has the potential to be applicable to 10-30\% of M dwarfs during their
early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, caused by long-lived, corotating clumps of cool
plasma \cite{CollierCameron1989,Donati2000,Dunstone2006,Skelly2008}.
Such clumps are forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
magnetic field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.
However, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether they are related to CPVs.
A circumstantial argument that they might is that CPVs do respond to
sudden magnetic field changes: the otherwise long-lived eclipse
features often disappear following stellar flares
\cite{Stauffer2017,Bouma2024}.

@@

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  Similar to
other CPVs \cite{Bouma2024}, the photometric shape of
TIC~141146667 evolved following the 2022 discovery
data, while nonetheless remaining complex.  In February 2024, the
average photometric signal showed a gradual brightening over 45\% of
the period, followed by a complex eclipse-like feature spanning 55\%
of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.
Its W-shape is suggestive of eclipse geometries seen in forward models
of warped plasma tori \cite{Townsend2008}.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
sharp transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to 
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.  The usual interpretation for
such variability comes from a loose analogy to
quiescent solar prominences, which are cool
condensations of plasma in the solar corona that can last days to
weeks \cite{VialEngvold2015}.
In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic or centrifugal force
capable of sustaining them.  However for stars with magnetospheric
radii $R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a plasma parcel can have a local minimum
outside $R_{\rm c}$, enabling the material to be sustained for
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
Characteristic densities and masses of these
clumps are $n \sim 10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see
Supplementary Methods Section~\ref{subsec:model}), a similar density
to solar prominences, but ten to one hundred times more massive.  This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} since such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios
in which the circumstellar material is made only of dust are also
ruled out.  While dust may be present, to explain the H$\alpha$
emission the circumstellar clumps must include plasma with a
significant population of hydrogen atoms in the $n$=3 excited state.
While this plasma is undoubtedly sculpted by the star's magnetic
field, it could plausibly originate from three sites: the star, an old
and undetected disk, or outgassing rocky bodies.  This latter
possibility would render CPVs as extrasolar analogs of the Jupiter-Io
plasma torus (CITE).

@@

The material's composition -- either pure plasma, or else a dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical have
previously suggested chromaticity consistent with dust \cite{Tanimoto2020,Gunther2022,Koen2023},
a gray opacity source such as electron scattering in a plasma
transiting over starspots could also produce chromatic features
\cite{Rackham2018}.  The composition and size distribution of any dust
that is present could be most easily resolved by measuring the
extinction curve for one or more CPVs from 1-10\,$\mu$m.  A
dust composition similar to debris from rocky bodies seen around white
dwarfs \cite{Reach2009} would indicate a rocky-body origin.  A 
composition closer to the ISM would be indicative of condensed dust in
an M dwarf wind, similar to that formed in the environments of more
evolved stars \cite{Marigo2008}.

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
(magneto)-hydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008} and to explore the effects of
non-dipolar field contributions.
In particular, dynamo simulations of fully-convective M dwarfs have
suggested the possibility that global-scale mean fields can be
confined to a single hemisphere \cite{Brown2020}; such fields would
yield accumulations of circumstellar material quite different
from those that have previously been explored \cite{Townsend2008}.
Observationally, spectropolarimetry has the potential to
assess both the field strength and topology.  An independent
probe could also be connected the recent work \cite{Kaur2024} showing
that CPVs are variable radio emitters with emission components
that can be both persistent, and also short-lived and highly
polarized.  In particular, detecting radio emission
from the electron cyclotron maser instability would
provide a direct measurement of the field strength at the site of the
emitting region.

@@

{\it Effective temperature, radius, and mass}
We adopt the results from the spectral energy distribution fitting
exercise described by \cite{Bouma2024}.
In brief, this approach used
\texttt{astroARIADNE}...
\cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012} assuming the
\cite{Asplund2009} solar abundances, and the
\cite{Barber2006} water line lists.  
We used broadband magnitudes from Gaia DR2, APASS, 2MASS, SDSS, and WISE
  $W1$ and $W2$.  (explain)...

%%%%%%%%
%\texttt{astroARIADNE} compares the measured broadband flux
%measurements against pre-computed model grids, and by default fits for
%six parameters: $\{ T_{\rm eff}, R_\star, A_{\rm V}, \log g, [{\rm
%Fe/H}], d \}$.  The distance  prior is drawn from
%\citet{2021AJ....161..147B}.  The surface gravity and metallicity are
%generally unconstrained.  Given our selection criteria for the stars,
%we assumed the following priors for the temperature, stellar size, and
%extinction:
%\begin{align}
%  T_{\rm eff} / {\rm K}    &\sim \mathcal{U}(2000, 8000), \\
%  R_\star / R_\odot  &\sim \mathcal{U}(0.1, 1.5), \\
%  A_{\rm V} / {\rm mag}    &\sim \mathcal{U}(0, 0.2),
%\end{align}
%for \deleted{$\mathcal{N}$ the Gaussian and }$\mathcal{U}$ the uniform
%distribution\deleted{s, and $\mathcal{T}_{\rm N}(\mu, \sigma, a, b)$ a
%truncated normal distribution with mean $\mu$, standard deviation
%$\sigma$, and lower and upper bounds $a$ and $b$}.  We validated our
%chosen upper bound on $A_{\rm V}$ using a 2MASS color-color diagram.
%Finally, using \texttt{Dynesty} \citep{2020MNRAS.493.3132S}, we
%sampled the posterior probability assuming the default Gaussian
%likelihood, and set a stopping threshold of ${\rm d}\log \mathcal{Z} <
%0.01$, where $\mathcal{Z}$ denotes the evidence.

{\it SED: No evidence for disk}
...
